The abundance of Bullet-Groups in $\Lambda$CDM

### 2014-03-20 paper/ms.tex
abstract

\title{The abundance of Bullet-Groups in $\Lambda$CDM}
\author{J. G. Fern\'andez-Trincado$^{1,2,3}$, J. E. Forero-Romero$^1$,
  T. Verdugo$^3$, G. Foex$^4$ and V. Motta$^4$} 
\affil{$^1$ Departamento de F\'{i}sica, Universidad de los Andes,
  Cra. 1 No. 18A-10, Edificio Ip, Bogot\'a, Colombia\\ 
       $^2$ Institute Utinam, CNRS UMR6213, Universit\'e de
  Franche-Comt\'e, OSU THETA de Franche-Comt\'e-Bourgogne,
  Besan\c{c}on, France\\ 
       $^3$ Centro de Investigaciones de Astronom\'ia, AP 264,
  M\'erida 5101-A, Venezuela\\
  $^4$ Instituto de F\'isica y Astronom\'ia, Universidad de
  Valpara\'iso, Avda. Gran Breta\~na 1111, Playa Ancha, Valpara\'iso
  2360102, Chile  
}

\begin{abstract}

We estimate the expected distribution of displacements between the two
dominant dark matter density peaks and between baryons and dark matter
in halos simulated within a full cosmological context. We use as a benchmark the
observation of \bullg, which is the lowest mass system observed so far
featuring a bi-modal dark matter distribution with a dislocated 
baryonic component. We extend previous computational results by
studying two halo samples: groups and clusters. We find that $(50 \pm 10)\%$
of the dark matter halos with circular velocities in the range
$300\kms$ to $700\kms$ (groups) show multimodal morphologies with
displacements between the dark matter clumps equal or larger than
$186 \pm 30$\hkpc as observed in \bullg. For dark matter halos with
circular velocities larger than $700\kms$ (clusters) this fraction
rises to $(70 \pm 10)\%$. Using the same simulation we estimate the
dark matter-to-baryon spatial separation and find that $0.1$ to
$1.0\%$ of the groups should present separations equal or
larger than $87\pm 14$\hkpc corresponding to our observational
benchmark; for clusters this fraction rises to $(7\pm 3)\%$, consistent
with previous studies of dark matter to baryon separations. In terms
of absolute numbers densities we find that the expected value for
bullet groups is tree times larger than the expected value for bullet
groups. The extension of the theoretical predictions and observational
results towards low mass bi-modal dark matter configurations opens up
the possibility for a new statistical test of  $\Lambda$CDM.  
\end{abstract}

\begin{keywords}
{cosmology: theory -- dark matter} 
\end{keywords}

@@

The Bullet Cluster (\bulla) provided a new kind of observational
evidence for the existence of dark matter
\citep{Markevitch2004,Clowe2006}. Quantifying the displacement between
dark matter and the dominant baryonic component (hot X-ray emitting gas)
has been used to test the Cold Dark Matter (CDM)
paradigm itself by quantifying the substructure velocity required to
produce such displacement \citep{Hayashi2006, Springel2007,
  Thompson2012}, the displacemeent between the dominant dark matter
and baryonic component \citep{ForeroRomero2010} to estimate the
expected abundance of such events in a $\Lambda$CDM Universe, putting
constraints on the dark matter particle self-interaction cross section
and even explore possible extensions to the concordance cosmological model
\citep{Farrar2007,Lee2010,Lee2012}.

@@

Considering that the statistics for the cluster sample are dominated
by objects in the mass range of the \bullg we make a comparison
against this sample. The table shows that $\sim 40\%$ of the 
clusters are expected to have large DM-DM displacements, this is a
factor of $\sim 2$ larger than the observational estimate by
\citep{Foex2013}. However, roughly $\sim 7\%$ of these systems present a
displacement equal or larger than the observed in the \bullg, a
fraction that is compatible with the $1/13\sim 0.07$ fraction in the
observational sample.

This rough comparison hints that it is possible to find a large fraction
of multi-modal groups with a DM-baryon displacement in a $\Lambda$CDM
cosmology and that our estimates for the relative number of bullet
systems with respect to multi-modal systems is compatible with
observations.

A proper comparison must take into account all the observational
uncertainties, biases, mixes between our two populations (groups \&
clusters at different redshifts) to derive a stronger prediction from
the simulation, something that is beyond the scope of this Letter.

>>>>

\title{The abundance of Bullet-Groups in $\Lambda$CDM}
\author{J. G. Fern\'andez-Trincado$^{1,2,3}$, J. E. Forero-Romero$^1$,
  T. Verdugo$^3$, G. Foex$^4$ and V. Motta$^4$} 
\affil{$^1$ Departamento de F\'{i}sica, Universidad de los Andes,
  Cra. 1 No. 18A-10, Edificio Ip, Bogot\'a, Colombia\\ 
       $^2$ Institute Utinam, CNRS UMR6213, Universit\'e de
  Franche-Comt\'e, OSU THETA de Franche-Comt\'e-Bourgogne,
  Besan\c{c}on, France\\ 
       $^3$ Centro de Investigaciones de Astronom\'ia, AP 264,
  M\'erida 5101-A, Venezuela\\
  $^4$ Instituto de F\'isica y Astronom\'ia, Universidad de
  Valpara\'iso, Avda. Gran Breta\~na 1111, Playa Ancha, Valpara\'iso
  2360102, Chile  
}
\email{jfernandezt87@gmail.com}
\email{je.forero@uniandes.edu.co}
\begin{abstract}

We estimate the expected distribution of displacements between the two
dominant dark matter density peaks and between baryons and dark matter
in halos simulated within a full cosmological context in $\Lambda$CDM.
We use as a benchmark the observation of \bullg, which is the lowest mass system observed so far
featuring a bi-modal dark matter distribution with a dislocated
baryonic component. We extend to lower masses previous computational results by
studying two halo samples: groups and clusters. We find that $(50 \pm
10)\%$ of the dark matter halos with circular velocities in the range
$300\kms$ to $700\kms$ (groups) show multimodal morphologies with
displacements between the dark matter clumps equal or larger than
$186 \pm 30$\hkpc as observed in \bullg. For dark matter halos with
circular velocities larger than $700\kms$ (clusters) this fraction
rises to $(70 \pm 10)\%$. Using the same simulation we estimate the
dark matter-to-baryon spatial separation and find that $0.1$ to
$1.0\%$ of the groups should present separations equal or
larger than $87\pm 14$\hkpc corresponding to our observational
benchmark; for clusters this fraction rises to $(7\pm 3)\%$, consistent
with previous studies of dark matter to baryon separations. In terms
of absolute number densities we find that the expected value for
groups similar to \bullg\ is tree times larger than the expected value
for clusters. This extension of the theoretical predictions and
observational results towards low mass bi-modal dark matter
configurations opens up the possibility for a new statistical test of
$\Lambda$CDM.   
\end{abstract}

\keywords{dark matter, galaxies: clusters: individual (SL2S
  J08544-0121), galaxies: interactions, methods: numerical}

@@

The Bullet Cluster (\bulla) provided a new kind of observational
evidence for the existence of dark matter
\citep{Markevitch2004,Clowe2006}. Since then it has been used to test the Cold
Dark Matter (CDM) paradigm itself by quantifying different aspects
such as the expected displacement between the dominant dark matter and
baryonic component in a $\Lambda$CDM Universe \citep{ForeroRomero2010}
or the substructure velocity required to produce such displacement \citep{Hayashi2006, Springel2007,
  Thompson2012}. It has also been used to constrain the dark matter
particle self-interaction cross section and to explore possible
extensions to the concordance cosmological model
\citep{Farrar2007,Lee2010,Lee2012}.

@@

Considering that the statistics for the cluster sample are dominated
by objects in the mass range of the \bullg we make a comparison
against this sample. Table \ref{table:numbers} shows that $\sim 40\%$ of the 
clusters are expected to have large DM-DM displacements, this is a
factor of $\sim 2$ larger than the observational estimate by
\citep{Foex2013}. However, roughly $\sim 7\%$ of these systems present a
displacement equal or larger than the observed in the \bullg, a
fraction that is compatible with the $1/13\sim 0.07$ fraction in the
observational sample.

This rough comparison  that our estimates for the relative number of
bullet systems with respect to multi-modal systems is compatible with
observations. A proper comparison must take into account all the observational
uncertainties, biases, mixes between our two populations (groups \&
clusters at different redshifts) to derive a stronger prediction from
the simulation, something that is beyond the scope of this Letter.

Neverthless, comparing the group and cluster samples we expect to have
a number density of bullets similar to \bullg that is three times
larger for groups than for clusters.
